---
title: GO-Term Analysis
date: "`r Sys.Date()`"
output:
  html_notebook:
    code_folding: hide
    fig_width: 12
    fig_height: 6
params:
  dat: ''
  output_path: '.'
  save_excel: TRUE
  significance_cutoff: 0.05
  simplify_ontologies: TRUE
  do_gse: TRUE
  use_background: TRUE
  store_r_objects: TRUE
  simplify_cutoff: 0.7
---

**Don't Panic**.

There is a lot of stuff in this document, just go through it step by step and try not not deal with everything at once.
Here is what you should do.

1. Check the Data Overview

We need to attach [EntrezID](https://www.ncbi.nlm.nih.gov/gene/)'s to each gene to conduct a GO-term or GSEA analysis.
However, for some genes this might fail, because no data base so far contains all known genes and transcripts.
Please check the section "Regulated genes that do not map to a EntrezID" in the *Data Overview* tab if there are genes that are important to you and that have no EntrezID assigned.

2. Check the Volcano plot.

Does the Volcano plot like you would expect?

3. Check GO-terms (*GO-Term All*)

The GO-term overrepresentation analysis uses all significant genes of your data set.
Check the figures and see if you find something interesting.
If there are too many GO-terms or they are too redundant, please check *GO-Term Simplified*.

4. Optional: Check *GSE GO Analysis*

A GSE uses _all_ detected genes and their fold change, rather than only the significant ones.
**This you should only do if there is absolutely nothing of interest in your GO-term analysis**.
You can also use this if you have a very subtle treatment and you expect very small genetic changes.

## {.tabset}

### Data Overview

Settings

<details>

```{r setup, echo = FALSE, message = TRUE, warning = TRUE, results = 'asis'}

# Retreive parameter settings

rmd_params <- list()

rmd_params$dat <- params$dat
rmd_params$output_path <- params$output_path
rmd_params$save_excel <- params$save_excel
rmd_params$significance_cutoff <- params$significance_cutoff
rmd_params$simplify_ontologies <- params$simplify_ontologies
rmd_params$do_gse <- params$do_gse
rmd_params$use_background <- params$use_background
rmd_params$store_r_objects <- params$store_r_objects
rmd_params$simplify_cutoff <- params$simplify_cutoff

library(magrittr)
suppressWarnings(library(mygo))
suppressWarnings(library(enrichplot))
suppressWarnings(library(org.Mm.eg.db))
suppressWarnings(library(AnnotationDbi))
suppressWarnings(library(stats4))
suppressWarnings(library(BiocGenerics))
suppressWarnings(library(parallel))
suppressWarnings(library(rmyknife))

# Print settings as table
tibble::tibble(
  name = c("output_path", "save_excel", "significance_cutoff", "simplify_ontologies", "do_gse", "use_background", "store_r_objects", "simplify_cutoff"),
  setting = c(rmd_params$output_path, rmd_params$save_excel, rmd_params$significance_cutoff, rmd_params$simplify_ontologies, rmd_params$do_gse, rmd_params$use_background, rmd_params$store_r_objects, rmd_params$simplify_cutoff)
) %>%
  knitr::kable()

# Summary for parameter file
message("Data table head")
rmd_params$dat %>%
  head() %>%
  knitr::kable()

if (!('q_value' %in% colnames(rmd_params$dat))) {
  stop('Column `q_value` not found in data frame')
} else if (!('fc' %in% colnames(rmd_params$dat))) {
  stop('Column `fc` not found in data frame')
}

```

</details>

```{r attach_entrez_id, results = 'asis'}

# Filter significant genes
if (!('EntrezID' %in% colnames(rmd_params$dat))) {
  warning('EntrezID not available in data frame. Attaching it manually')
  dat <- rmd_params$dat %>%
    ensembl_to_entrez(ensembl_id_name = "ensembl_gene_id", keep_only_rows_with_entrez = TRUE, drop_duplicates = TRUE) %>%
    attach_gene_symbol_from_entrez()
  
  # Get regulated genes that don't have a EntrezID
  # https://github.sf.mpg.de/bruening-lab/2018-01-Weiyi-Differential-Gene-Expression-Analysis/blob/master/Analysis_DESeq.Rmd
  message("Regulated genes that do not map to a EntrezID")
  rmd_params$dat %>%
    dplyr::filter(q_value <= rmd_params$significance_cutoff) %>%
    dplyr::anti_join(dat %>% dplyr::filter(q_value <= rmd_params$significance_cutoff), by = "ensembl_gene_id") %>%
    # Attach gene name from Biomart entry
    rmyknife::attach_biomart(attributes = "external_gene_name", verbose = FALSE) %>%
    knitr::kable() %>%
    print()
  
  message("Check for genes with duplicated entries. We will consider the ones with the lowest p-value")

  dat %<>% add_is_duplicated("EntrezID")
  dat %>%
    dplyr::filter(EntrezID_is_duplicated == T) %>%
    knitr::kable() %>%
    print()
  
  dat %<>%
    # For duplicated Entrez IDs only keep the ones with the lowest p-value
    dplyr::group_by(EntrezID) %>%
    dplyr::summarize(q_value = min(q_value)) %>%
    # Join back with the original data set in which we will also remove entries that have the same EntrwzID, q-value and FC
    # Because it can happen that a EntrezID maps to different Ensembl-IDs
    dplyr::inner_join(dat %>% dplyr::distinct(EntrezID, q_value, fc, .keep_all = TRUE))
  # Finally check if we still have problems regarding multiple entries
  .temp_multiple_entries <- dat %>% add_is_duplicated('EntrezID') %>% dplyr::filter(EntrezID_is_duplicated == T)
  if (nrow(.temp_multiple_entries) > 0) {
    warning("Duplicates per EntrezID, check input data")
    .temp_multiple_entries %>% kable()
  }
  # Remove obsolete rows
  dat %<>% dplyr::select(-EntrezID_is_duplicated)
}
# Get names vector from EntrezIDs
# dat %<>% attach_gene_symbol_from_entrez()
# Named double vector of fold changes named after the gene symbol
fc_symbol <- dat %>%
  .$fc %>%
  as.double() %>%
  `names<-`(dat$Symbol)
```

### Volcano Plot

Dataset summary

- We count `r nrow(dat)` genes.
- `r dplyr::filter(dat, q_value <= 0.05) %>% nrow()` genes are significantly regulated.
- `r dplyr::filter(dat, q_value <= 0.05 & fc > 0) %>% nrow()` genes are significantly up-regulated.
- `r dplyr::filter(dat, q_value <= 0.05 & fc < 0) %>% nrow()` genes are significantly down-regulated.

```{r volcano_plot_ggplot}

dat %>%
  # Draw Volcano plot
  ggplot2::ggplot(
    ggplot2::aes(
      x = fc,
      y = -log10(q_value),
      color = (q_value <= 0.05) %>% ifelse(., "significant", "not significant")
    )
  ) +
    ggplot2::geom_point(
      alpha = 0.3,
      size = 0.5
    ) +
    ggplot2::scale_color_manual(values = c("grey", "blue")) +
    ggplot2::xlab(expression(log[2](fc))) + 
    ggplot2::ylab(expression(-log[10](adjusted~p~value))) +
    ggplot2::labs(colour = "Significance") +
    ggthemes::theme_tufte(base_family = "sans") +
    ggrepel::geom_text_repel(
      data = . %>%
        dplyr::arrange(q_value) %>%
        head(10),
      mapping = ggplot2::aes(label = Symbol),
      size = 3
    ) +
    ggplot2::geom_hline(
      yintercept = -log10(0.05),
      linetype = "dotted"
    )

```

### GO-Term All

Gene-Ontology (GO)-term analysis is also called a GO-over represenation test.

We use the over-representation test implemented in clusterProfiler and [described here](https://bioconductor.org/packages/release/bioc/vignettes/clusterProfiler/inst/doc/clusterProfiler.html#go-over-representation-test)

The over-representation test is carried out internally with the DOSE package and is [described here](https://bioconductor.org/packages/release/bioc/vignettes/DOSE/inst/doc/enrichmentAnalysis.html#over-representation-analysis)

For details on the hypergeometric distribution, refer to [GO::TermFinder—open source software for accessing Gene Ontology information and finding significantly enriched Gene Ontology terms associated with a list of genes](https://academic.oup.com/bioinformatics/article/20/18/3710/202612)

  - Hypergeometric distribution is a distribution modelling normally distributed probabilities (e.g. red, blue and green marbles in a Urn) -- [Wiki](https://de.wikipedia.org/wiki/Hypergeometrische_Verteilung)
  - Freely spoken, DOSE determines whether any terms annotated by the list of differentially expressed genes is greater than the one expected by chance in a term


```{r get_go_and_gse}

# Expect a data set containing q_values and Entrez/Ensembl IDs
dat_goterms <- dat %>% get_go_all_ontologies(
  significance_cutoff = rmd_params$significance_cutoff,
  use_background = rmd_params$use_background
)
# Get simplified ontologies if required
if (rmd_params$simplify_ontologies) {
  dat_goterms_simplified <- simplify_ontologies(
    ontologies = dat_goterms,
    cutoff = rmd_params$simplify_cutoff
  )
}
if (rmd_params$do_gse) {
  dat_gse_go <- dat %>% get_gse_all_ontologies()
  dat_gse_kegg <- dat %>% get_kegg()
} else {
  dat_gse_go <- list()
  dat_gse_go$Biological_Process <- tibble::tibble()
  dat_gse_go$Molecular_Function <- tibble::tibble()
  dat_gse_go$Cellular_Components <- tibble::tibble()
  dat_gse_kegg <- list()
  dat_gse_kegg$kegg <- tibble::tibble()
}
```

```{r plot_go}

# dat_goterms %>% plot_all_ontologies(fc_symbol)

```

#### Biological Process

Here is the list of all regulated GO-terms sorted by significance.

```{r bp_print_table}
#' Helper function, takes clusterProfiler GO-term result and prints it as DT widget
#' @export
#' @import DT dplyr tibble magrittr
#' @param dat clusterProfiler GO-term result
print_goterm_as_datatable <- function(dat) {
  dat %>%
    tibble::as_tibble() %>%
    dplyr::select(-geneID, -pvalue, -p.adjust) %>%
    DT::datatable(
      extensions = 'Buttons',
      options = list(
        dom = 'Bfrtip',
        buttons = c('copy', 'csv', 'excel', 'pdf', 'print')
      ),
      filter = "top"
    ) %>%
    return()
}
dat_goterms$Biological_Process %>%
  print_goterm_as_datatable()
  
```

The Enrich Map plot shows overlapping genes between GO-terms as grey edged between each node.
Each node represents a regulated GO-term.
Only the top 50 regulated GO-terms are plotted here.

```{r bp_emap, fig.width = 14, fig.height = 14}

dat_goterms$Biological_Process %>%
  mygo::emap_plot('Enrich Map Plot Plot', 50)

```

The following scatter plot displays all regulated GO-terms w.r.t total GO-term gene count (GO-term size) on the x-axis and percentage of significant genes (how many of the GO-term genes are significantly regulated) on the y-axis.
The top 15 regulated GO-terms are labeled.

```{r bp_overlap_scatterplot, fig.width = 14, fig.height = 10}

dat_goterms$Biological_Process %>%
  mygo::attach_goterm_genecount(species = "MUS") %>%
  mygo::overlap_scatterplot()

```

#### Molecular Function

```{r mf_print_table}
dat_goterms$Molecular_Function %>%
  print_goterm_as_datatable()
```
```{r mf_emap, fig.width = 14, fig.height = 14}
dat_goterms$Molecular_Function %>%
  mygo::emap_plot('Enrich Map Plot Plot', 50)
```
```{r mf_overlap_scatterplot, fig.width = 14, fig.height = 10}
dat_goterms$Molecular_Function %>%
  mygo::attach_goterm_genecount(species = "MUS") %>%
  mygo::overlap_scatterplot()
```

#### Cellular Components

```{r cc_print_table}
dat_goterms$Cellular_Components %>%
  print_goterm_as_datatable()
```
```{r cc_emap, fig.width = 14, fig.height = 14}
dat_goterms$Cellular_Components %>%
  mygo::emap_plot('Enrich Map Plot Plot', 50)
```
```{r cc_overlap_scatterplot, fig.width = 14, fig.height = 10}
dat_goterms$Cellular_Components %>%
  mygo::attach_goterm_genecount(species = "MUS") %>%
  mygo::overlap_scatterplot()
```


### GO-Term Simplified

ClusterProfiler allows usage of the `simplify` method that removes redundant GO terms. For details, refer to [this blog post](https://guangchuangyu.github.io/2015/10/use-simplify-to-remove-redundancy-of-enriched-go-terms/).

```{r print_go_ontologies}
if (rmd_params$simplify_ontologies) {
  dat_goterms_simplified %>% plot_all_ontologies(fc_symbol)
} else {
  print("Simplified ontologies deactivated.")
}
```

### GSE GO Analysis

* The GO-overrepresentation test can miss subtle changes in terms, for example when all genes in a term are up-regulated, but not enough the render each gene differentially regulated
* The GSEA uses all genes to detect changes where all genes in a predefined set change in a small but coordinated way
* Implemented in clusterProfiler and [described here](https://bioconductor.org/packages/release/bioc/vignettes/clusterProfiler/inst/doc/clusterProfiler.html#go-gene-set-enrichment-analysis)
* The original paper is [Gene set enrichment analysis: A knowledge-based approach for interpreting genome-wide expression profiles](http://www.pnas.org/content/102/43/15545.long)

```{r gse_all}
if (rmd_params$do_gse) {
  dat_gse_go %>% plot_all_ontologies(fc_symbol)
} else {
  print("GSE analysis disabled")
}
```

### GSE KEGG Pathways

Here we apply the GSE analysis to KEGG pathways.

```{r gse_kegg_all}
if (rmd_params$do_gse) {
  dat_gse_kegg %>% plot_all_ontologies(fc_symbol)
} else {
  print("GSE analysis disabled")
}
```

```{r export_result}
# Create dummy variable for for simplified gene ontologies if we did not calculate it
if (!rmd_params$simplify_ontologies) {
  dat_goterms_simplified <- data.frame()
}

# Store clusterProfiler objects to disk
if (rmd_params$store_r_objects) {
  dat_goterms %>%
    saveRDS(file = file.path(rmd_params$output_path, "dat_goterms.rds"))
  
  if (rmd_params$do_gse) {
    dat_gse_go %>%
      saveRDS(file = file.path(rmd_params$output_path, "dat_gse_go.rds"))
    dat_gse_kegg %>%
      saveRDS(file = file.path(rmd_params$output_path, "dat_gse_kegg.rds"))
  }
  if (rmd_params$simplify_ontologies) {
    dat_goterms_simplified %>%
      saveRDS(file = file.path(rmd_params$output_path, "dat_goterms_simplified.rds"))
  }
}

# Export result
if (rmd_params$save_excel) {
  export_go_terms_to_excel(
    go_ontologies        = dat_goterms,
    go_ontologies_simple = dat_goterms_simplified,
    gse_ontologies       = dat_gse_go,
    kegg_ontologies      = dat_gse_kegg,
    path                 = paste0(rmd_params$output_path, '/go_terms.xlsx')
  )
}
```
